HSF Community White Paper on Data Organization, Management and Access (DOMA)

### 2018-03-28 CWP/papers/HSF-CWP-2017-04_doma/latex/hsf-cwp-dm.tex
CWP DOMA: language fixes and other cleanup by Bo
- Imported from Overleaf on April 5, 10 am CET

\abstract{To be done.}

@@

\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
 & & HSF-CWP-2017-04 \\
 & & March 11, 2018 \\ % use \date or hardwire e.g. December 15, 2017
 & & \\
\end{tabular*}
\vspace{2.0cm}

@@

The coming decade will bring HEP to the exabyte scale. The expected data
volumes of the HL-LHC experiments, the DUNE experiment, and even non-HEP
experiments such as SKA (Square Kilometre Array) will all enter this
regime. In devising experimental computing models for this era, many
factors have to be taken into account. ~In particular, the increasing
availability of very high-speed networks, which may reduce the need for
CPU and data co-location, need to be examined. Such networks may allow
for more extensive use of data access over the wide-area network (WAN),
which may provide failover capabilities, global and federated data
namespaces, and will have an impact on data caching. Shifts in the data
presentation/analysis models such as a potential move to event-based
data streaming from the more traditional dataset-based or file-based
data access, will be particularly important for optimizing the
utilization of opportunistic computing cycles on HPC facilities,
commercial cloud resources, and campus clusters, and can potentially
resolve currently limiting factors such as job eviction.

Planned analysis of HL-LHC data and data analysis from other large
international collaborations will need to adopt a distributed computing
model due to its scope. The data management systems that will enable
this type of analysis will need to provide cost-based optimization of
the data locations and delivery mechanisms.

@@

The LHC experiments currently provision and manage about an Exabyte of
storage, approximately half of which is archival, and half is
traditional disk storage. The storage requirements per year are expected
to jump by a factor of 10 for the HL-LHC. This itself is faster than
projected Moore's Law gains and will present major challenges. Storage
will remain one of the visible cost drivers for HEP computing, however
the projected growth and cost of the computational resources needed to
analyze the data is also expected to grow even faster than the base
storage costs. The combination of storage and analysis computing costs
may restrict scientific output and potential physics reach of the
experiments. Thus, new techniques and algorithms are likely to be
required.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item \textbf{Big Data:} The HL-LHC era will significantly increase both the
  date rate and the data volume. The computing systems will need to
  handle this without significant cost increases and within evolving
  storage technology limitations.
\item \textbf{Dynamic Distributed Computing:} In addition, the significantly
  increased computational requirements for the HL-LHC era will also
  place new requirements on data. Specifically, the use of new types of
  compute resources (\emph{e.g.,} cloud, HPC) with different dynamic
  availability and characteristics are used will require more dynamic
  DOMA systems.
\item \textbf{New Applications:} New applications, such as machine learning
  training or high-rate data-query systems for analysis, will likely be
  employed to meet the computational constraints and to extend the
  physics reach of the HL-LHC. These new applications will place new
  requirements on how and where data is accessed and produced. For
  example, specific applications (e.g. training for machine learning)
  may require use of specialized processor resources such as GPUs,
  placing further requirements on data.
\end{enumerate}

The rapid increase in recent years of data-intensive problems in both
the commercial world and in the rest of the research world also provides
a number of opportunities.

@@

The original LHC computing models (circa 2005) were built up from the
simpler models used before distributed computing was a central part of
HEP computing. This allowed for a reasonably clean separation between
three different aspects of interacting with data: organization,
management and access.

\begin{itemize}
\item \textbf{Data Organization:} This is essentially how data is structured
  as it is written. Most data is written in flat files, in ROOT format,
  typically with a column-wise organization of the data. The records
  corresponding to these columns are compressed. The internal details of
  this organization are typically visible only to individual software
  applications.
\item \textbf{Data Management:} The key challenge here was the transition to
  the use of distributed computing in the form of the grid. The
  experiments developed dedicated data transfer and placement systems,
  along with catalogs, to move data between computing centers. To first
  order, the computing models were rather static: data was placed at
  sites and the relevant compute jobs were sent to the right locations.
  Applications might interact with catalogs or, at times, the workflow
  management systems does this on behalf of the applications.
\item \textbf{Data Access:} Various protocols are used for direct reads
  (rfio, dcap, xrootd, etc.) with a given computer center and/or
  explicit local stage-in and caching for read by jobs. Application
  access may use different protocols than those used by the data
  transfers between sites.
\end{itemize}

Before the LHC turn-on and in the first years of the LHC, these three
areas were, to first order, optimized independently. Many of the
challenges were in the area of ''Data Management (DM)'' as the Worldwide
LHC Computing Grid was commissioned. As the LHC computing matured
through Run1 and Run2, the interest has turned to optimizations spanning
these three areas. For example, the recent use of ``Data Federations''
mixes up Data Management and Access. As we will see below, some of the
foreseen opportunities towards HL-LHC may require global optimizations.

Thus in this document we take a broader view than traditional ``DM'',
and consider the combination of ``Data Organization, Management and
Access (DOMA)'') together. We believe that this full picture of data
needs in HEP will provide important opportunities for efficiency and
scalability as we enter the many-Exabyte era.

@@

  technologies can provide. In particular the internal structure of the

@@

  streaming.) and the actual analysis calculations being performed can
  dramatically affect the CPU to IO ratios. These factors make this

@@

  data transfers between WAN endpoints (e.g. 10-100Gbps.) This

@@

  the network performance. Satisfying both requirements is necessary to
  provide an efficient point to point data flow.

@@

In addition to the characteristics described above, HEP workflows in
each category exhibit a high degree of parallelism, which allow the work
to map readily into both high throughput computing (HTC) and high
performance computing (HPC) environments. These domains span wide
dynamic ranges in their requirements for concurrent data access. If the
experiments of the HL-LHC, DUNE and other large international projects
are to utilize these facilities, they will need the ability to integrate
with multiple storage technologies. The storage technologies will need
to be tuned or matched to the corresponding computing resources that the
experiments are targeting their workflows to execute on.

These storage models will also need to be designed and optimized for
cost, in the context of the scientific workflows that they are enabling.
Simultaneously, these storage models must remain flexible enough to
support evolving physical storage options similar to the the shifts that
have occurred in recent years between mRAM, SSDs, shingled disks, tapes
and other technologies. ~

@@

The HEP community has long established patterns that have been used for
data organization and have shaped the prevailing data access and storage
models that are in use today. These established patterns have focused
around sets of event records organized into discrete collections through
traditional file based storage. There are, however, potential gains that
can be made by exploring alternative data organization models and the
corresponding access and storage models that they can enable. These
alternative designs should be evaluated for the HL-LHC era and, if found
to be advantageous, implemented by the community. A shift in designs may
have the potential to reduce storage costs, shift network utilisation,
and result in more efficient use of compute resources.

There are two main avenues that these organizational changes can follow,
each with different impacts on the analysis methodologies. First, we
should reconsider the structure of the data, which can be row oriented,
column oriented, have varying levels of granularity, represent larger or
smaller aggregations of objects etc, boosting the efficiency of
particular applications.

Secondly, we should reconsider the granularity at which data management
systems operate. This is currently aggregations of events (files, or
larger sets such as blocks or datasets), and could shift towards being
the event (or even sub-event), potentially reducing storage volumes. In
that case, other types of ``container'' beyond files could be considered
(e.g. objects).

Design decisions will be subject to competing requirements from data
management and access optimisation, and will have consequences
throughout the system. We have to tackle a global optimization problem,
taking into account all use cases and their respective frequencies to
further optimise application performance for a given resource
investment.

@@

The structure of data has to be adapted to applications' requirements.
At one limit, the data could be organized simply as events, the other
extreme is to organize the data as columns which represent sub-event
level objects such as particles. Event-based organization favors
applications like event displays while column-based organization favors
large statistical analysis. Our current system of data tiers in files
draws intermediate (and overlapping) rectangles between these two
extremes. Finding the right balance will be an important factor in
efficient data analysis.

@@

Decisions on granularity and structure have the potential of broadening
the scope of technologies that can be used to store the data; for
example, exploiting object stores (such as AWS S3) as event stores. In
addition, this would support parallel reading and writing of events as
event objects from applications. This could allow applications to better
scale to large numbers of CPU core where now these applications
concurrency is limited by the requirement to sequentially write events
to files. Currently, the efficiency of of reading and writing events
from object stores is sensitive to the size of the event. The smaller
the event the worse is the efficiency.. Some of this may be mitigated by
creating event bundles (i.e. several events in one object). The
evolution of the ATLAS Event Service may provide some of the answers and
should be supported. As a parallel effort, the ROOT team is also looking
into increasing concurrency within the ROOT framework. This effort
should also be supported as it may have more immediate success in the
near term.

Commercial providers offer innovative storage services (``cloud
storage'') which may fit certain use cases well. The relevance of these
solutions to computing models must be understood, not only through
technical metrics such as performance, but with an understanding of the
underlying cost structure and the risks associated with such
procurements. Reliance on commercially procured cloud storage as a core
component of data organizational models would require a potential
fundamental shift in user policies and discussion with funding agencies.

\subsection{Optimizing Data Management
}\label{optimizing-data-management}

Data Management policy will directly affect data access. For example,
the decision to distribute multiple enriched data samples optimized for
different use cases would improve data access and thus compute
efficiency, at the expense of larger volumes transferred and stored.
Similarly, the unit of data management could be changed, and become the
event, which promises to reduce storage volumes and promote the use of
opportunistic resources through enabling pre-emptibility in jobs. This
would have knock-on effects on data catalogues, data distribution and
storage systems, and would affect data access. Where an experiment
positions itself amongst all these possibilities will be dictated by its
policy. Thus the potential gains in terms of efficient use of resources
should be realised through data management systems which allow this
policy to be expressed and implemented.

@@

A key requirement in devising data catalog models is understanding the
required granularity of data to be cataloged and what data for an
experiment needs to be centrally cataloged or tracked, and for what
purpose (accounting, metadata, location). Analysis access to the data,
for instance, may benefit from being below the event level (e.g. physics
object). Any such catalogs may have to be external to the primary data
store (e.g., relational databases that work with metadata in the primary
data store). If cataloging requirements for production and
reconstruction differ considerably from analysis, an approach with
multiple, complementary catalogs can also be considered. The cataloging
schemes chosen will be tightly coupled with the data organization
models. For example, in the case where data organization is object-based
or content-addressable storage is used, the catalogs need only a single
handle to have a fully descriptive location of a given event or other
object being sought.

@@

Data delivery models can become a limiting factor in performance,
especially as they heavily depend on shared resources such storage
systems and the network. Caching has to be considered to avoid data
delivery bottlenecks and unnecessary transfers. At present, HEP data
delivery models are designed solely with the traditional sequential
processing paradigm used in reconstruction in mind. Computing models
where the only form of persistent storage is on tape and all, or nearly
all, disk-based storage is considered as a caching layer should also be
considered, in which case work on optimising workflows for cache
efficiency and exploiting data popularity metrics would be valuable.

@@

A useful analogy for data delivery methods might be content delivery
network services (CDNs) such as the one offered by Akamai. Without CDNs,
on-demand streaming of movies over the Internet or even the delivery of
the web sites of major news outlets like CNN or BBC would quickly
collapse. What is the equivalent of content delivery for accessing huge
datasets over distance, and for which a particular analysis might only
require a fraction of the dataset? CDNs use, to great benefit, highly
sophisticated, proprietary naming, caching, and placement techniques
that are customized to particular types of data and what is known about
their access patterns. Similarly intelligent techniques could be used to
place, compose, and deliver scientific data products that minimizes data
movement, latency, and other costs, taking advantage of data semantics
and information about the location of reusable intermediate results.
This is not unlike query planning in relational databases, except for
the scale and the different data type of scientific data: a query
optimizer might be able to recognize that data can be filtered remotely,
or combined from data products cached in the neighborhood instead of
having to fetch everything across continents. New approaches, like Named
Data Networking (NDN), could play a role: by allowing to name, at the
network level, the data you are looking for rather than its location, it
opens the paths for many optimizations behind the scenes based on
factors such as the client application location or the data type.

Currently, new data-caching strategies are being deployed (e.g., XRootD
caching proxy servers, server-less caching, ARC-style caching) and may
provide sufficient information on how such strategies can optimize
mammoth data delivery task for HL-LHC event reconstruction and analysis.
However, without a global view on how these strategies are used, they
will likely have a limited impact on improving data delivery efficiency.

@@

Data analysis is currently tied in with ROOT-based formats. In many
currently-used paradigms, physicists consider all events at an
equivalent level of detail and in the format offering the highest level
of detail that needs to be considered in an analysis. ~However, not
every event considered in analysis requires the same level of detail.
One consideration to improve event access throughput is to design event
tiers with different abstractions, and thus data sizes. All events can
be considered at a lighter-weight tier while events of interest only can
be accessed with a more information-rich tier.

@@

\section{Performance : Tools and
Metrics}\label{performance-tools-and-metrics}

As the size of available datasets grows, we must allow for the
possibility of increased user-driven selection and dataset generation.
In terms of reducing the strain on data access, techniques such as data
augmentation can prove useful. Cluster-computing frameworks (e.g.,
Spark) can also be leveraged to reduce the time of dataset reduction and
access. It will be necessary to study and classify IO patterns in
applications used in order to understand how data access methods can be
optimized across prevalent access patterns. Pattern data can then be
used in algorithms resulting from active research in computer science
that study payoffs between storage and cpu in various computing models.

A first correlated analysis of the IO patterns of different
computational tasks has already started at the CERN computing center and
within several experiments. In particular the combination of
infrastructure and experiment information (e.g. detailed knowledge of
hardware capabilities and network topology, per job CPU and storage
metrics) may allow us to define more meaningful performance metrics than
the traditionally used raw data rate and CPU utilization which both show
their limitations. These metrics should enable a joint quantitative
comparison of different storage and computing approaches across
individual system boundaries and hence lead to a more effective resource
investment.

@@

The projected event complexity of data from future LHC runs and from
high resolution liquid argon detectors will require advanced
reconstruction algorithms and analysis tools to understand. The
precursors of these tools, in the form of new machine learning paradigms
and pattern recognition algorithms, already are proving to be drivers
for the CPU needs of the HEP community. As these techniques continue to
grow and blossom, they will place new requirements on the computational
resources that need to be leveraged by all of HEP. The storage systems
that are developed, and the data management techniques that are employed
will need to directly support this wide range of computational
facilities, and will need to be matched to the changes in the
computational work, so as not to impede the improvements that they are
bringing.

Storage will remain one of the visible cost drivers for HEP computing,
but the projected growth and cost of the computational resources needed
to analyze the data are expected to grow faster than the base storage
costs. The combination of storage and analysis computing costs may
restrict scientific output and potential physics reach of the
experiments. There must be R\&D efforts in data management on how to
minimize the impact of the data access and storage model on the overall
cost of doing scientific analysis. This R\&D should include an
optimization of both the capital costs of storage, as well as the
potential impacts the storage systems can have on the CPU requirements
for the experiments and their costs.

The ability to leverage new storage technologies as they become
available into existing data delivery models is a challenge that we must
be prepared for. New storage systems will present new interfaces and new
behaviour. A key decision in their successful exploitation will be
whether to encapsulate this within a more familiar service or whether to
present the new interface directly to applications which will thus have
to be adapted. As discussed in the preceding sections, much of this
change can be aided by active R\&D into our own IO patterns; an approach
which has not yet been adopted widely by the field.

@@

On the hardware side, R\&D is needed in alternative approaches to data
archiving to determine the possible cost/performance tradeoffs.
Currently, tape is extensively used to hold data that cannot be
economically made available online. While the data is still accessible,
it comes with a high latency penalty; limiting possible analysis. We
suggest investigating either separate direct access-based archives (e.g.
disk or optical) or new models that overlay online direct access volumes
with archive space. This is especially relevant when access latency is
proportional to storage density. Either approach would need to also
evaluate reliability risks and the effort needed to provide data
stability.

Cost reductions in maintenance and operation of the storage
infrastructure can be realised through convergence of the major
experiments and resource providers on shared solutions. This does not
necessarily mean promoting a monoculture, as different solutions will be
adapted to certain major classes of use-case, type of site or funding
environment. Indeed, there will always be a judgment to make on the
desirability of using a variety of specialised systems, or abstracting
the commonalities through a more limited but common interface (the SRM
story illustrates this point). Reduced costs and improved sustainability
will be further promoted by extending these concepts of convergence
beyond HEP and into the other large-scale scientific endeavours that
will share the infrastructure in the coming decade. Efforts must be made
as early as possible, during the formative design phases of such
projects, to create the necessary links.

Finally, any and all changes undertaken must not make the ease of access
to data any worse than it is under current computing models. We must
also be prepared to accept the fact that the best possible solution may
require significant changes in the way data is handled and analyzed.
What is clear is that what is being done today will not scale to the
needs of HL LHC.

@@

>>>>

\abstract{Without significant changes to data organization, management, and access (DOMA), HEP experiments will find scientific output limited by how fast data can be accessed and digested by computational resources. In this white paper we discuss challenges in DOMA that HEP experiments such as the HL-LHC will face as well as potential ways to address them. A research and development timeline to assess these changes is also proposed.}

@@

\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
 & & HSF-CWP-2017-04 \\
 & & April 3, 2018 \\ % use \date or hardwire e.g. December 15, 2017
 & & \\
\end{tabular*}
\vspace{2.0cm}

@@

The coming decade will bring HEP to the exabyte (EB) scale with the expected data
volumes of the HL-LHC experiments, the DUNE experiment, and even non-HEP
experiments such as the Square Kilometre Array (SKA) entering this
regime. In devising experimental computing models for this era, many
factors have to be taken into account. In particular, the increasing
availability of very high-speed networks, which may reduce the need for
% * <briedel@uchicago.edu> 2018-04-04T17:59:27.886Z:
%
% > very high-speed networks
% What is "very high-speed" here? 100G? 1000G?
%
% ^ <briedel@uchicago.edu> 2018-04-04T17:59:40.128Z.
CPU and data co-location, need to be examined. Such networks may allow
for more extensive use of data access over the wide-area network (WAN),
which may provide fail-over capabilities, global and federated data
namespaces, and will have an impact on data caching. Shifts in the data
presentation and analysis models, such as a potential move to or
sub-event-based or event-based data streaming from the more
% * <briedel@uchicago.edu> 2018-04-04T18:00:17.686Z:
%
% > sub-event-based or event-based data
% Added "sub-event-based" because we talk about it below.
%
% ^ <briedel@uchicago.edu> 2018-04-04T18:00:36.656Z.
traditional dataset-based or file-based data access, will be
particularly important for optimizing the utilization of
opportunistic computing cycles on HPC facilities, commercial cloud resources,
and campus clusters, and can potentially resolve currently
limiting factors such as job eviction.
% * <briedel@uchicago.edu> 2018-04-04T18:01:55.490Z:
%
% >  job eviction
% Using "job evictions" as a reason to move to different DOMA model seems like an odd choice. Shouldn't it be something like "scheduling difficulties"
%
% ^.

Planned analysis of HL-LHC data and data analysis from other large
% * <briedel@uchicago.edu> 2018-04-04T18:09:21.988Z:
%
% > large
% XENON1T has shown that the collaboration doesnt have to be "large" to require DOMA
%
% ^.
international collaborations will need to adopt a distributed computing
model due to its scope. The data management systems that will enable
this type of analysis will need to provide cost-based optimization of
the data locations and delivery mechanisms.

@@

The LHC experiments currently collectively provision and manage about an EB of
storage of which approximately half is archival and half is
traditional disk storage. The annual storage requirements are expected
to jump by a factor of 10 for the HL-LHC. This increase is faster than
projected Moore's Law gains and will present major challenges. Storage
% * <briedel@uchicago.edu> 2018-04-05T01:05:55.296Z:
%
% > Moore's Law gains
% This is only a CPU. What about storage technology? Currently we are adding about 2 TB/year in disk capacity. It is unlikely will be at 100 TB/disk in 2023. We may have O(10) - O(100) TB SSDs  at a reasonable cost, but we can't be certain. What does the future of tape look like?
%
% ^.
will remain one of the visible cost drivers for HEP computing with
the projected growth and cost of the computational resources needed to
analyze the data expected to grow even faster than the base
storage costs. The combination of storage and analysis computing costs
may restrict scientific output and potential physics reach of the
experiments. Thus, new techniques and algorithms are likely to be
required.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item \textbf{Big Data:} The HL-LHC era will significantly increase both expected
  date rates and data volumes. Computing systems will need to
  handle this without significant cost increases and within evolving
  storage technology limitations.
\item \textbf{Dynamic Distributed Computing:} In addition, the significantly
  increased computational requirements for the HL-LHC era will also
  place new requirements on data. Specifically, the use of new types of
  compute resources (e.g., cloud, HPC) with different dynamic
  availability and characteristics will require more dynamic
  DOMA systems.
\item \textbf{New Applications:} New applications, such as machine learning
  training or high-rate data-query systems for analysis, will likely be
  employed to meet the computational constraints and to extend the
  physics reach of the HL-LHC. These new applications will place new
  requirements on how and where data is accessed and produced. For
  example, specific applications (e.g., training for machine learning)
  may require use of specialized processor resources, such as GPUs,
% * <briedel@uchicago.edu> 2018-04-04T18:14:36.074Z:
%
% > ,  placing further requirements on data.
% Not sure this is needed.
%
% ^.
  placing further requirements on data.
\end{enumerate}

The rapid increase in recent years of data-intensive problems in both
the commercial world and in the rest of the research world also provides
a number of opportunities and solution to tackle these challenges.

@@

The original LHC computing models (developed circa 2005) were built up from
simpler models used before distributed computing was a central part of
HEP computing. This allowed for a reasonably clean separation between
three different aspects of interacting with data: organization,
management and access.

\begin{itemize}
\item \textbf{Data Organization:} This is essentially how data is structured
  as it is written. Most data is written in flat files, in the ROOT format, and
  typically with a column-wise organization of the data. The records
  corresponding to these columns are compressed. The internal details of
  this organization are typically visible only to individual software
  applications.
\item \textbf{Data Management:} The key challenge here was the transition to
  the use of distributed computing in the form of the grid. The
  experiments developed dedicated data catalog,  transfer and placement systems,
  to move data between computing centers. To first
  order, the computing models were rather static: data was placed at
  sites and the relevant compute jobs were sent to the right locations.
  Applications might interact with catalogs or, at times, the workflow
  management systems does this on behalf of the applications.
\item \textbf{Data Access:} Various protocols are used for direct reads
  (rfio, dcap, xrootd, etc.) with a given computer center and/or
  explicit local stage-in and caching for read by jobs. Application
  access may use different protocols than those used by the data
  transfers between sites.
\end{itemize}

Before the LHC turn-on and in the first years of data taking, these three
areas were, to first order, optimized independently. Many of the
challenges were in the area of ''Data Management (DM)'' as the Worldwide
LHC Computing Grid was commissioned. As the LHC experiments' computing matured
through Run1 and Run2, interest has turned to optimizations spanning
these three areas. For example, the recent use of ``Data Federations''
must consider together both Data Management and Access. As we will see below, some of the
foreseen opportunities towards HL-LHC may require global optimizations.

Thus, in this document we take a broader view than traditional ``DM'',
and consider the combination of ``Data Organization, Management and
Access (DOMA)'' together. We believe that this full picture of data
needs in HEP will provide important opportunities for efficiency and
scalability as we enter the many-EB era.

@@

  technologies can have. In particular, the internal structure of the

@@

  streaming) while the actual analysis calculations being performed can
  dramatically affect the CPU-to-IO ratios. These factors make this

@@

  data transfers between WAN endpoints (e.g. 10-100~Gbps). This

@@

  network performance. Satisfying both requirements is necessary to
  provide efficient point-to-point data flow.

@@

In addition to the characteristics described above, HEP workflows in
each category exhibit a high degree of parallelism, which allow the work
to map readily into both high throughput computing (HTC) and high
performance computing (HPC) environments. These domains span wide
dynamic ranges in their requirements for concurrent data access. If the
experiments of the HL-LHC, DUNE and other large international projects
are to utilize these facilities, they will need the ability to integrate
with multiple storage technologies. The storage technologies will need
to be tuned or matched to the corresponding computing resources that the
experiments are targeting.

These storage models will also need to be designed and optimized for
cost, in the context of the scientific workflows that they are enabling.
Simultaneously, these storage models must remain flexible enough to
support evolving physical storage options similar to the the shifts that
have occurred in recent years between mRAM, SSDs, shingled disks, tapes,
and other technologies.

@@

The HEP community has long-established data organization patterns that have
shaped the prevailing data access and storage
models that are in use today. These established patterns are centered
around sets of event records organized into discrete collections through
traditional file-based storage. There are, however, potential gains that
can be made by exploring alternative data organization models and the
corresponding access and storage models that they can enable. These
alternative designs should be evaluated for the HL-LHC era and, if found
to be advantageous, implemented by the community. A shift in designs may
have the potential to reduce storage costs, shift network utilization,
and result in more efficient use of compute resources.

There are two main avenues that these organizational changes can follow,
each with different impacts on the analysis methodologies. First, we
should reconsider the structure of the data, which can be row-oriented,
column-oriented, have varying levels of granularity, or represent larger or
smaller aggregations of objects, boosting the efficiency of
particular applications.

Secondly, we should reconsider the granularity at which data management
systems operate. This is currently at aggregations of events (files, or
larger sets, such as blocks or datasets), and could shift towards being
the event (or even sub-event), potentially reducing storage volumes. In
that case, other types of ``containers'' beyond files could be considered
(e.g. objects).

Design decisions will be subject to competing requirements from data
management and access optimization, and will have consequences
throughout the system. We have to tackle a global optimization problem,
taking into account all use cases and their respective frequencies to
further optimize application performance for a given resource
investment.

@@

The structure of data has to be adapted to application requirements.
At one limit, the data could be organized simply as events, while the other
extreme is to organize the data as columns which represent sub-event
level objects such as particles. Event-based organization favors
applications like event displays while column-based organization favors
large statistical analysis. Our current system of data tiers in files
draws intermediate (and overlapping) rectangles between these two
extremes. Finding the right balance will be an important factor in
efficient data analysis.

@@

Decisions on granularity and structure have the potential of broadening
the scope of technologies that can be used to store the data; for
example, exploiting object stores (such as AWS S3 or Ceph) as event stores. Such an approach
would support parallel reading and writing of events as
event objects from applications. This could allow applications to better
scale to large numbers of CPU cores whereas in current models application
concurrency is limited by the requirement to sequentially write events
to files. Currently, the efficiency of of reading and writing events
from object stores is sensitive to the size of the event---
the smaller the event, the worse the efficiency is. Some of these potential inefficiencies
may be mitigated by creating event bundles (i.e. several events in one object). The
evolution of the ATLAS Event Service may provide some of the answers and
should be supported. As a parallel effort, the ROOT team is also looking
into increasing concurrency within the ROOT framework. This effort
should also be supported as it may have more immediate success in the
near term.

Commercial providers offer innovative storage services (``cloud
storage'') which may fit certain use cases well. The relevance of these
solutions to HEP computing models must be understood, not only through
technical metrics such as performance, but with an understanding of the
underlying cost structure and the risks associated with such
procurements. Reliance on commercially procured cloud storage as a core
component of data organizational models would require a potential
fundamental shift in user policies and discussion with funding agencies.

\subsection{Optimizing Data Management}\label{optimizing-data-management}

Data Management policy will directly affect data access. For example,
the decision to distribute multiple enriched data-samples optimized for
different use cases would improve data access and thus compute
efficiency, at the expense of larger volumes transferred and stored.
Similarly, the unit of data management could be changed, and become the
event, which promises to reduce storage volumes and promote the use of
opportunistic resources through enabling pre-emptibility in jobs. This
would have cascading effects on data catalogs, data distribution and
storage systems, and would affect data access. Where an experiment
positions itself amongst all these possibilities will be dictated by its
policy. Thus the potential gains in terms of efficient use of resources
should be realized through data management systems which allow this
policy to be expressed and implemented.

@@

A key requirement in devising data catalog models is understanding the
required granularity of data to be cataloged and what data for an
experiment needs to be centrally cataloged or tracked, and for what
purpose (accounting, metadata, location). Analysis access to the data,
for instance, may benefit from being below the event level (\emph{e.g.}, physics
object). Any such catalogs may have to be external to the primary data
store (\emph{e.g.}, relational databases that work with metadata in the primary
data store). If cataloging requirements for production and
reconstruction differ considerably from analysis, an approach with
multiple, complementary catalogs can also be considered. The cataloging
schemes chosen will be tightly coupled with the data organization
models. For example, in the case where data organization is object-based
or content-addressable storage is used, the catalogs need only a single
% * <briedel@uchicago.edu> 2018-04-05T02:04:26.292Z:
%
% > content-addressable
% This comes a little out of nowhere. Where is this discussed before?
%
% ^.
handle to have a fully descriptive location of a given event or other
object being sought.
% * <briedel@uchicago.edu> 2018-04-05T02:05:03.704Z:
%
% > or other object
% This is the first time "other object" is mentioned. What other object? TBranches? It would be better to call this sub-event data like done before.
%
% ^.

@@

Data delivery models can become a limiting factor in performance,
especially as they heavily depend on shared resources such storage
systems and the network. Caching has to be considered to avoid data
delivery bottlenecks and unnecessary transfers. At present, HEP data
delivery models are designed solely with the traditional sequential
processing paradigm used in reconstruction in mind. Computing models
where the only form of persistent storage is on tape and all, or nearly
all, disk-based storage is considered as a caching layer should also be
considered, in which case work on optimizing workflows for cache
efficiency and exploiting data popularity metrics would be valuable.

@@

A useful analogy for data delivery methods might be content delivery
network services (CDNs) such as the one offered by Akamai. Without CDNs,
% * <briedel@uchicago.edu> 2018-04-05T02:08:36.870Z:
%
% > Akamai.
% Google also offers one
%
% ^.
on-demand streaming of movies over the Internet or even the delivery of
the web sites of major news outlets like CNN or BBC would quickly
collapse. What is the equivalent of content delivery for accessing huge
datasets over distance where a particular analysis might only
require a fraction of the dataset? CDNs use, to great benefit, highly
sophisticated proprietary naming, caching, and placement techniques
that are customized to particular types of data and what is known about
their access patterns. Similarly intelligent techniques could be used to
place, compose, and deliver scientific data products that minimizes data
movement, latency, and other costs, taking advantage of data semantics
and information about the location of reusable intermediate results.
This is not unlike query planning in relational databases, except for
the scale and the different data type of scientific data: a query
optimizer might be able to recognize that data can be filtered remotely,
or combined from data products cached in the neighborhood instead of
having to fetch everything across continents. New approaches, like Named
Data Networking (NDN), could play a role. NDNs allow the naming, at the
network level, of the data you are looking for rather than its location, which
opens the paths for many optimizations behind the scenes based on
factors such as the client application location or the data type.

Currently, new data-caching strategies are being deployed (e.g., XRootD
caching proxy servers, server-less caching, ARC-style caching) and may
% * <briedel@uchicago.edu> 2018-04-05T02:09:39.626Z:
%
% > server-less caching
% In-memory caching?
%
% ^.
provide sufficient information on how such strategies can optimize
mammoth data delivery task for HL-LHC event reconstruction and analysis.
However, without a global view on how these strategies are used, they
will likely have a limited impact on improving data delivery efficiency.

@@

Data analysis in most of HEP is currently tied to ROOT-based formats. In many
currently-used paradigms, physicists consider all events at an
equivalent level of detail and in the format offering the highest level
of detail that needs to be considered in an analysis. ~However, not
every event considered in analysis requires the same level of detail.
One consideration to improve event access throughput is to design event
tiers with different abstractions, and thus data sizes. All events can
be considered at a lighter-weight tier while events of interest only can
be accessed with a more information-rich tier.

@@

\section{Performance: Tools and
Metrics}\label{performance-tools-and-metrics}

As the size of available datasets grows, we must allow for the
possibility of increased user-driven selection and dataset generation.
In terms of reducing the strain on data access mechanisms, techniques such as data
augmentation can prove useful. Cluster-computing frameworks (e.g.,
Spark) can also be leveraged to reduce the time of dataset reduction and
access. It will be necessary to study and classify I/O patterns in
applications used in order to understand how data access methods can be
optimized across prevalent access patterns. Pattern data can then be
used in algorithms resulting from active research in computer science
that study trade-offs between storage and CPU in various computing models.

A first correlated analysis of the I/O patterns of different
computational tasks has already started at the CERN Data Centre and
within several experiments. In particular, the combination of
infrastructure and experiment information (e.g. detailed knowledge of
hardware capabilities and network topology, per job CPU and storage
metrics) may allow us to define more meaningful performance metrics than
the traditionally-used raw data rate and CPU utilization which both have
their limitations. These metrics should enable a joint quantitative
comparison of different storage and computing approaches across
individual system boundaries and thus lead to a more effective resource
investment.

@@

The projected event complexity of data from future LHC runs and from
high resolution liquid argon detectors will require advanced
reconstruction algorithms and analysis tools to understand. The
precursors of these tools, in the form of new machine learning paradigms
and pattern recognition algorithms, already are proving to be drivers
for the CPU needs of the HEP community. As these techniques continue to
gain traction in our field and evolve, they will place new requirements on the computational
resources that need to be leveraged by all of HEP. The storage systems
that are developed and the data management techniques that are employed
will need to directly support this wide range of computational
facilities, and will need to be matched to the changes in the
computational work so as not to impede the improvements that they are
bringing.

Storage will remain one of the visible cost drivers for HEP computing,
but the projected growth and cost of the computational resources needed
to analyze the data are expected to grow faster than the base storage
costs. The combination of storage and analysis computing costs may
restrict scientific output and potential physics reach of the
experiments. There must be R\&D efforts in data management on how to
minimize the impact of the data access and storage model on the overall
cost of doing scientific analysis. This R\&D should include an
optimization of both the capital costs of storage as well as the
potential impacts the storage systems can have on the CPU requirements
for the experiments and their costs.

The ability to leverage new storage technologies, as they become
available, into existing data delivery models is a challenge that we must
be prepared for. New storage systems will present new interfaces and new
behavior. A key decision in their successful exploitation will be
whether to encapsulate this within a more familiar service or whether to
present the new interface directly to applications which will thus have
to be adapted. As discussed in the preceding sections, much of this
change can be aided by active R\&D into our own IO patterns; an approach
which has not yet been adopted widely by the field.

@@

On the hardware side, R\&D is needed in alternative approaches to data
archiving to determine the possible cost/performance trade-offs.
Currently, tape is extensively used to hold data that cannot economically be made available online. While the tape-stored data is still accessible,
it comes with a high latency penalty and thus limiting possible analysis. We suggest investigating either separate direct-access-based archives (e.g.
disk or optical) or new models that overlay online direct-access volumes
with archive space. This is especially relevant when access latency is
proportional to storage density. Either approach would need to also
evaluate reliability risks and the effort needed to provide data
stability.

Cost reductions in maintenance and operation of the storage
infrastructure can be realized through convergence of the major
experiments and resource providers on shared solutions. This does not
necessarily mean promoting a monoculture, as different solutions will be
adapted to certain major classes of use-case, type of site or funding
environment. Indeed, there will always be a judgment to make on the
desirability of using a variety of specialized systems, or abstracting
the commonalities through a more limited but common interface (the SRM
story illustrates this point). Reduced costs and improved sustainability
will be further promoted by extending these concepts of convergence
beyond HEP and into the other large-scale scientific endeavors that
will share the infrastructure in the coming decade. Efforts must be made
as early as possible, during the formative design phases of such
projects, to create the necessary links.

Finally, any and all changes undertaken must not make the ease of access
to data any worse than it is under current computing models. We must
also be prepared to accept the fact that the best possible solution may
require significant changes in the way data is handled and analyzed.
What is clear is that what is being done today will not scale to the
needs of HL-LHC experiments.

@@

\subsection{Data Access and Data Access Patterns}\label{access-patterns}
The data access pattern should inform the the use of storage media and data location. Proposed areas of inquiry are:
